---
title: "Food_Expenses_Quarto"
format: html
editor: visual
---

## Setup

```{r}
vroom :: vroom('FoodExpenses.txt')
```

## Main Conclusions - Questions

Describe the primary conclusions about the relationship between income and food
expenditures. What relationships did you find?

INSERT TEXT HERE

The advisers specifically hypothesize that a “healthy” restaurant economy should see
increases of about \$0.50 or more per week for each \$1000 increase in income. Based on the
data you have, do you think this is a healthy economy?

```{r}

```

Advisers are expecting incomes to generally shift upward in time (income generally increases
by 3-4% each year). Make sure you provide some predictions on what an upward shift in
income over the next 5 years or so could do to food expenditures (including the variance not
just the amount spent on EatingOut).

```{r}

```

## Modeling Details

Justify your hiring by showing that standard regression models won’t work for this situation so
they needed to hire you and should hire you again.

Describe this new “advanced” model you are using in mathematical detail and how you used it
to arrive at the conclusions you came to in the following section. In doing so, these advisers
Food Expenditures Consulting
Problem Background
n = 523
don’t know linear algebra so try to avoid that if possible.

Justify (i.e validate) that this model is a trustworthy model for the answers you wrote in the
previous section.
