---
title: "Heteroskedastic Model"
author: "Matthew and Corbin"
format: pdf
editor: visual
---

```{r}
#| warnings: false

library(tidyverse)
library(gamlss)

```

```{r}

food <- vroom::vroom("FoodExpenses.txt")

```

```{r}

# make heteroskedastic mdl using gamlss
het_food <- gamlss(formula = EatingOut ~ Income, sigma.formula = ~., data = food)
```

# checking assumptions

```{r}

car::avPlots(lm(EatingOut ~ Income, data = food))

```

```{r}

std_resids <- residuals(het_food, what = "z-scores")
```

```{r}
hist(std_resids, freq = F, xlim = c(-4,4))
par(new = T)
curve(dnorm(x), xlim = c(-4,4))

ks.test(std_resids,"pnorm", mean = 0, sd = 1)
```

```{r}

lm_df <- tibble(
  fitted = fitted(het_food),
  resids = std_resids
)

bp_pval <- lmtest::bptest(std_resids~Income, data = food)$p.value
bp_str <- paste0('BP Test p-value: ',round(bp_pval,3))

ggplot(data = lm_df, aes(x = fitted, y = resids))+
  geom_point(alpha = 1, size = 1)+
  ylim(-3,3)+
  geom_hline(yintercept = 0, color = 'red')+
  annotate(
      "text",
      x = 50, y = 3,
      label = bp_str,
      size = 4,
      color = "blue"
    )

```

```{r}
r_2 <- cor(food$EatingOut, fitted(het_food))^2
```

#### Advisers are expecting incomes to generally shift upward in time (income generally increases by 3-4% each year). Make sure you provide some predictions on what an upward shift in income over the next 5 years or so could do to food expenditures (including the variance not just the amount spent on EatingOut)

```{r}
#im not sure this is right. I'm thinking of a perfecnt increase as increase by one.035 but Im not so sure about it. It doesnt seem to connect perfectlty to a 1 unit increase. I think it depends on how many units we have to start

# 3-4% by year is essentiall 1.035 * B or B*.035 increase
beta_inc <- coef(het_food, what = "mu")["Income"]

theta_inc <- coef(het_food, what = "sigma")["Income"]

eat_increase <- .035 * beta_inc
eat_sd <- .035 * theta_inc

print("Yearly increase")
print(eat_increase)

print("sd increase")
print(eat_sd)

```

#### Describe this new “advanced” model you are using in mathematical detail and how you used it
to arrive at the conclusions you came to in the following section. In doing so, these advisers
donʼt know linear algebra so try to avoid that if possible.

Our advanced model uses a very similar technique to the traditional linear regression model to calculate a prediction for the expected Eating Out level at a certain income. The difference comes in allowing a change in the expected variance in Eating Out levels when income changes. Traditionally, it is expected that the amounts that predictions vary from the true value to be constant regardless of different Income amounts. In this model, we recognize that the amount that eating out habits varies increases as income increases. This allows us to better calculate confidence intervals and hypothesis tests.

I CAN ASSUME THEY UNDERSTAND LOG(SIGMA_2). I NEED TO PROVIDE EACH COEFFICIENT
